Towards caring 15-minute neighbourhoods

### 2024-06-21 data/data-raw/SOM-share/Session 11 - Self Organizaing Maps IV.Rmd
Started my edits, pushing back to Anastasia to incorporate the discussion of today's meeting with Lucia and Lea.

Load the libraries used in this notebook:
```{r load-libraries, include = FALSE}
library(readxl) # Read Excel Files
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(rgdal) # Bindings for the 'Geospatial' Data Abstraction Library
library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
library(rgeos) # Interface to Geometry Engine - Open Source ('GEOS')
library(dplyr) # A Grammar of Data Manipulation
library(tmap) # Thematic Maps
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(rpart) # Recursive Partitioning and Regression Trees
library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
```

>>>>

Load the libraries used in this notebook:
```{r load-libraries, include = FALSE}
library(readxl) # Read Excel Files
library(tidyverse) # Easily Install and Load the 'Tidyverse'
#library(rgdal) # Bindings for the 'Geospatial' Data Abstraction Library
library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
#library(rgeos) # Interface to Geometry Engine - Open Source ('GEOS')
library(dplyr) # A Grammar of Data Manipulation
library(tmap) # Thematic Maps
library(SOMbrero) # SOM Bound to Realize Euclidean and Relational Outputs
library(rpart) # Recursive Partitioning and Regression Trees
library(rpart.plot) # Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
```

### 2024-06-21 paper/Manuscript.qmd
Started my edits, pushing back to Anastasia to incorporate the discussion of today's meeting with Lucia and Lea.

---
title: "Towards caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  docx
  # elsevier-pdf:
  #   keep-tex: true
  #   include-in-header:
  #     text: |
  #       \usepackage{graphicx}
  #       \usepackage{unicode-math}
  #       \usepackage{hyperref}
  #       \def\tightlist{}
  #       \usepackage{setspace}
  #       \doublespacing
  #       \usepackage{lineno}
  #       \linenumbers
  #   journal:
  #     name: 7th International Conference on Women and Gender in Transportation (WGiT)
  #     formatting: preprint
  #     model: 3p
  #     cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-Minute City", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

The 15-Minute City is a normative conceptualisation gaining ground in urban planning: it frames neighbourhoods as responsive to human needs and environmental sensibilities, where most daily necessities can be reached within a 15-minute walk or bike ride [@Allam2022]. As a related tool, accessibility measures (the ease of reaching opportunities) are increasingly important amongst transport planners aiming to foster just and sustainable cities [@vale2023accessibility]. Both the 15-Minute City and accessibility measures are flexible enough to consider all destination types holistically however, gendered examinations have been lacking in the literature. For instance, accessibility analyses have historically focused on employment-centric and discretionary travel, types of travel more frequent for working-age and higher-income men.

To counter this masculinist bias, this study investigates a way to gender-mainstream the 15-Minute City through a care lens [e.g., @lawWomenTransportNew1999; @cresswell_gendered_2008; @Oestergaard1992; @levy1991towards; @levy2013travel] supported by the Mobility of Care framework (coined by @sanchezdemadariagaMobilityCareIntroducing2013). Mobility of Care emphasizes the importance of travel to unpaid work (care trips) in contrast to the better-studied travel to employment and leisure. While all three trip types (work, care, and leisure) are essential, care trips are often relatively shorter-distance, proximate to residential/work/school, and comprise approximately 30% of adults’ daily trips [@sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory; @mejia2021mobility]; fitting well within the 15-Minute City conceptualisation.

Our study provides an empirical example that maps the 15-Minute City onto the Mobility of Care framework. Specifically, it identifies which areas in Hamilton, Canada are 'caring 15-minute neighbourhoods'. To do so, a database of care destinations is created using secondary data. In this database, care destinations include all places associated with sustaining household tasks needed for the reproduction of life namely: shopping (e.g., groceries), errands (e.g., libraries), health (e.g., dentist), and caring for dependents (e.g., schools). This database is used to estimate the number and mix of care-destinations that can be reached within a 15-minute walk- and cycling- sheds from census centroids. Typologies are generated to illustrate which neighbourhoods can and to what degree facilitate 15-minute access to care.

Results indicate only a few neighbourhoods outside of the downtown core are 'care-complete', i.e., contain a sufficiently high mix of care destinations from all categories and sub-categories. However, some neighbourhoods are almost 'care-complete' and provide 15-minute access to some care categories. Our study frames these neighbourhoods on the continuum of 'caring' and in need of further intervention. The quantitative investigation conducted provides a high-level picture of what neighbourhoods (and their underlying land-use) are connected to transport infrastructure that can support reaching care-destinations.

Taken together, this study provides a theoretical bridge to connect 15-Minute Cities, accessibility analysis and Mobility of Care framework for the purpose of informing policy choice aimed to encourage just and sustainable mobility.

@@

Caring is one of the most fundamental activities in societies, which has allowed humanity to survive and evolve. It involves activities targeted to meet the physical, psychological, and emotional needs of others [@ilo_care_2018]. Yet, it is one of the most devalued, underpaid, and unequal activities worldwide. Traditionally, caring activities have remained on women’s shoulders. Though women today attain higher educational levels and tend to work out of home more than ever before, the bulk of care remains (at any life stage) heavily on women's shoulders. The 'Second Shift' remains mostly women’s work  in all industrial societies [@hayden_grand_2000; @hochschild_second_2012]. According to the @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care. This gender gap varies geographically [@RN5]. In Canada, adult women complete 1.9 hours, on average, more of unpaid care than men every day [@RN4].

This unequal share of caring responsibilities leads to gendered differences in: career development, profession selection, contract type, pay gap, and time poverty as recognized by different international organisations [@EIGE2016; @ilo_care_2018]. In fact, even early feminists in the United States observed that domestic labor was the most relevant basic cause of women’s inequality [@gordon_womans_1976; @hayden_grand_2000].

Care also has significant planning considerations. Indeed, researchers from various fields have highlighted that western modern industrial societies have imposed a physical separation of the private sphere from the public one, splitting production and reproduction, public and private spaces or simply work and home, concentrating on the economic outcome of being mobile [@cresswell_gendered_2008; @lawWomenTransportNew1999; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. The man with legal rights, paid labour public sphere and woman in the private sphere, dependent, with family duties, personal relations and unpaid labour [@tronto_toward_1990]. In this line, @cresswell_gendered_2008 described how masculinity was related to mobile and active, whereas femininity to static. As an example of this dichotomy, transport planning historically framed non-work/school trips as 'non-mandatory' activities, downgrading these activities in comparison to work. In more recent years, planners have begun to resist this spatial separation of the public and private sphere to promote for sustainable cities. Proximity planning has become more normalized, at popular example of which is the 15-minute city, in which residents can reach all necessary destinations within a 15-minute walk. The 15-minute city concept, however, has yet to directly focus on care.

@@

In Paris, the adoption of the 15-Minute city put into the spotlight this approach. They targeted six social functions that had to be easily accessible at any part of the city: housing, work, health care, groceries, education and leisure [@ville_de_paris_paris_2022]. In 2021, Ottawa also adopted a 15-Minutes approach in the new City of Ottawa Official Plan [@ville_dottawa_quartier_2021]. More recently, @moreno_introducing_2021 considers four dimensions on how the 15-minute city framework: density (in terms of people per sq.km); diversity (including mixed-land use and diversity of people; the temporal and spatial proximity to essential services (residential areas, work, commercial areas, education centres, health facilities and other basic institutions) and digitalisation.  In this respect, the feminist perspective of the cities of proximity is still underestimated. Only few examples exist that specifically target this issue [@gil_sola_choose_2022; @macintyre_her_2022]. Given the impact of caring activities on everyday life, the chrono-urbanism should specifically target the analysis of caring activities.

>>>>

---
title: "Towards caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  docx
  # elsevier-pdf:
  #   keep-tex: true
  #   include-in-header:
  #     text: |
  #       \usepackage{graphicx}
  #       \usepackage{unicode-math}
  #       \usepackage{hyperref}
  #       \def\tightlist{}
  #       \usepackage{setspace}
  #       \doublespacing
  #       \usepackage{lineno}
  #       \linenumbers
  #   journal:
  #     name: 7th International Conference on Women and Gender in Transportation (WGiT)
  #     formatting: preprint
  #     model: 3p
  #     cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

The 15-minute neighbourhood is a concept in planning that promotes urban spaces that are responsive to human needs and environmental sensibilities. These neighbuorhoods aim to accomplish this by ensuring that most daily necessities are not further away than a 15 minutes active trip [@Allam2022]. In practice, 15-minute neighborhoods are a form of normative accessibility, [@paez_measuring_2012], and therefore related to accessibility analysis, an increasingly important tool that planners use to foster just and sustainable cities [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. Both the 15-minute neighborhood and accessibility measures are flexible enough to consider all destination types holistically, however, gendered examinations have been lacking in the literature. For instance, accessibility analyses have historically been employment-centric even when adopting a gendered perspective [@paez_jobs_2024].

To address this gender bias, this study investigates the 15-minute neighbourhood through the lens of Mobility of Care [@@sanchezdemadariagaMobilityCareIntroducing2013]. This framework emphasizes the importance of travel related to unpaid work (care trips) in contrast to the better-studied travel to employment and leisure [@lawWomenTransportNew1999; @cresswell_gendered_2008; @Oestergaard1992; @levy1991towards; @levy2013travel]. While these trip types (work, care, and leisure) are essential, care trips are often relatively shorter-distance, proximate to residential/work/school, and comprise approximately 30% of adults’ daily trips [@sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory; @mejia2021mobility], thus fitting neatly within the concept of 15-minute neighbourhoods.

Our study casts the 15-minute neighbourhood onto the Mobility of Care framework. Empirically, we identify which areas in Hamilton, Canada, are 'caring 15-minute neighbourhoods'. To do so, a database of care destinations was compiled. In this database, care destinations include all places associated with sustaining household tasks needed for the reproduction of life namely: shopping (e.g., groceries), errands (e.g., libraries), health (e.g., dentist), and caring for dependents (e.g., schools). This database is used to estimate the number and mix of care-destinations that can be reached within a 15-minute walk-sheds from census centroids. We generate typologies to illustrate which neighbourhoods can and to what degree facilitate 15-minute access to care.

Our results indicate which 15-minute neighbourhoods are almost 'care-complete', that is, contain a sufficiently high mix of care destinations. Very few of these are found outside of the downtown core. Our study frames these neighbourhoods on a continuum of 'caring' and helps identify areas for policy intervention. Taken together, this study provides a theoretical bridge to connect and Mobility of Care, 15-minute neighbourhoods, and accessibility analysis. This should help to inform policy choices aimed at encouraging more just and sustainable mobility.

@@

Caring is one of the most fundamental activities in societies, since it consists of numerous activities that allow them to thrive. Caring activities are those that strive to meet the physical, psychological, and emotional needs of others [@ilo_care_2018]. Yet, it is one of the most unequal, undervalued, and even devalued activities worldwide. Traditionally, caring activities have been borne on women’s shoulders. Though women today attain higher educational levels and tend to work out of home more than ever before, the 'Second Shift' remains mostly women’s work  in all industrial societies [@hayden_grand_2000; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care, a gender gap that varies geographically [@RN5]. In Canada, adult women complete 1.9 hours, on average, more of unpaid care than men every day [@RN4].

This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In fact, early feminists in the United States had already observed that domestic labor was the single most relevant cause of women’s inequality in society [@gordon_womans_1976; @hayden_grand_2000].

Care also has important implications for spatial and transportation planning. Indeed, researchers from various fields have highlighted that western modern industrial societies have imposed a physical separation of the private sphere from the public one, splitting production and reproduction, public and private spaces or simply work and home, concentrating on the economic outcome of being mobile [@cresswell_gendered_2008; @lawWomenTransportNew1999; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. The man with legal rights, paid labour public sphere and woman in the private sphere, dependent, with family duties, personal relations and unpaid labour [@tronto_toward_1990]. In this line, @cresswell_gendered_2008 described how masculinity was related to mobile and active, whereas femininity to static. As an example of this dichotomy, transport planning historically framed non-work/school trips as 'non-mandatory' activities, downgrading these activities in comparison to work. In more recent years, planners have begun to resist this spatial separation of the public and private sphere to promote for sustainable cities. Proximity planning has become more normalized, at popular example of which is the 15-minute neighbourhood, in which residents can reach all necessary destinations within a 15-minute walk. The 15-minute neighbourhood concept, however, has yet to directly focus on care.

@@

In Paris, the adoption of the 15-minute neighbourhood put into the spotlight this approach. They targeted six social functions that had to be easily accessible at any part of the city: housing, work, health care, groceries, education and leisure [@ville_de_paris_paris_2022]. In 2021, Ottawa also adopted a 15-Minutes approach in the new City of Ottawa Official Plan [@ville_dottawa_quartier_2021]. More recently, @moreno_introducing_2021 considers four dimensions on how the 15-minute neighbourhood framework: density (in terms of people per sq.km); diversity (including mixed-land use and diversity of people; the temporal and spatial proximity to essential services (residential areas, work, commercial areas, education centres, health facilities and other basic institutions) and digitalisation.  In this respect, the feminist perspective of the cities of proximity is still underestimated. Only few examples exist that specifically target this issue [@gil_sola_choose_2022; @macintyre_her_2022]. Given the impact of caring activities on everyday life, the chrono-urbanism should specifically target the analysis of caring activities.
